Real-time resolution of short-read assembly graph using ONT long reads

### 2020-09-08 plos_latex_clean.tex
Lachlan revision 1

%% END MACROS SECTION
\usepackage[normalem]{ulem}
\newcommand\cancel[1]{\sout{#1}}
\newcommand\addnew[1]{\textcolor{red}{#1}}
%% npGraph
\newcommand{\npscarf}{$\mathtt{npScarf}$}
\newcommand{\npscarfg}{$\mathtt{npScarf\_wag}$}
\newcommand{\npreader}{$\mathtt{npReader}$}
\newcommand{\npanalysis}{$\mathtt{npAnalysis}$}
\newcommand{\npbarcode}{$\mathtt{npBarcode}$}
\newcommand{\npgraph}{$\mathtt{npGraph}$}
\newcommand{\canu}{$\mathtt{Canu}$}
\newcommand{\unicycler}{$\mathtt{Unicycler}$}
\newcommand{\spades}{$\mathtt{SPAdes}$}
\newcommand{\hspades}{$\mathtt{hybridSPAdes}$}
\newcommand{\albacore}{$\mathtt{Albacore}$}
\newcommand{\racon}{$\mathtt{Racon}$}
\newcommand{\metrichor}{$\mathtt{Metrichor}$}
\newcommand{\minimap}{$\mathtt{minimap2}$}
\newcommand{\miniasm}{$\mathtt{miniasm}$}
\newcommand{\bwa}{$\mathtt{BWA\text{-}MEM}$}

@@

One particular strength of ONT technology is the production of ultra long reads. 
This is complementary to the dominant short read sequencing technology Illumina which is cheaper and has higher per base read quality but is unable to resolve the complex regions of the genome due to its read length limitation. 
\addnew{Hybrid assembly with ONT reads provides the opportunity to use long reads to fully resolve existing microbial genome assemblies. However, in order achieve this in an effective manner it is necessary to avoid both under-sequencing (and thus not completely resolving the assembly) or over-sequencing (and thus incurring unnecessary costs and time-to -results).  In addition, the release of Read Until API provides the opportunity to selectively enrich parts of the genome, as has been implemented in customised targeted sequencing applications ~\cite{Payne2020}.   The combination of ReadUntil with  streaming hybrid assembly opens the possibility for further efficiency gains by targeting sequencing to unresolved regions of the genome. }

@@

The first step is to cluster the big anchors (longer than $10$Kbp and in/out degree less than 2) based on their \emph{kmer} coverage.
To achieve this, we applied DBSCAN clustering algorithm \cite{Ester96adensity-based}, which utilises a customised metric function to map contigs into a one-dimensional space.  In order to 
 define a customised metric which is \cancel{sample} \addnew{simple} and fast to calculate, we assumed that a single long contig itself consists of a Poisson distribution of \emph{k-mers} count with the mean approximated by the contig's coverage. 
%The rationale is to approximate a coverage value of an extremely long contig (which can be split into more than 10,000 \emph{k-mers}) with a sampled mean of a Poisson distribution (of \emph{k-mers} count). 
The metric is then determined by a distance function of two Poisson distributions based on Kullback-Leibler divergence (or relative entropy) between the Poisson distribution representing each contig\cite{Kullback1951information}.

@@

Due to false \addnew{positive} alignments from shorter contigs to the long reads, not all of the reported step nodes \cancel{are necessary to be appeared} \addnew{would appear} in the ultimate path resolved by the bridge. 
In most cases, the accumulated score of each step indicates its likelihood to be the true component of the final solution.
For that reason, a strategy similar to binary search\cancel{ing} is employed to find a path across 2 anchors of a bridge as shown in Algorithm~\ref{algo:binsearch}. \addnew{Briefly, it recursively looks for sub-paths connecting pairs of step nodes in-between 2 anchors and return a final continuous path with the highest score.}
\cancel{Before that, }We define Algorithm~\ref{algo:findpath} \addnew{as the \emph{base} sub-rountine} to \cancel{demonstrates the path finding algorithm for} \addnew{find paths between a \emph{connectable} pair of} two \addnew{step} nodes given their estimated distance. In which, function 
\[
\mathtt{shortestTree}(\overrightarrow{vertex},distance) : (V,Z) \rightarrow V^n
\]
from line 3 \cancel{of the algorithm's pseudo code} builds a shortest tree rooted from $\overrightarrow{v}$, following its direction until a distance of approximately $d$ (with a tolerance \cancel{regarding nanopore read} \addnew{for the nanopore} error rate) is reached. This task is implemented based on Dijkstra algorithm.
This tree is used on line 4 and in function $includedIn()$ on line 19 to filter out any node or edge with ending nodes that do not belong to the tree.

@@

% Results and Discussion can be combined.
\section*{Results}
\subsection*{Evaluation using synthetic data}
To evaluate the performance of the method, \npgraph{} 1.1 was tested along with \spades{}, \spades{} hybrid from version 3.13.1, ~\cite{AntipovKM2015}, \npscarf{} (japsa 1.7-02), and \unicycler{} version 0.4.6 using \unicycler{}`s synthetic data set~ \cite{Wick2017unicycler} . 
\addnew{}
The data set is a simulation of Illumina and MinION raw data, generated \emph{in silico} based on available microbial references.
We ran hybrid assembly methods using the entire nanopore data and the reciprocal results were evaluated by QUAST 5.0.2~\cite{Mikheenko2018quast5}.

@@

@@

\begin{figure}[!ht]
 \includegraphics[width=\textwidth]{images/FigS2.eps}
 \caption{Alignments of an \emph{Enterobacter~cloacae} reference genome to assembly sequences generated by  (a)~\cancel{\unicycler{}}\addnew{\npgraph{}} and (b)~\cancel{\npgraph{}}\addnew{\unicycler{}}. The former suggests a structural variant, the latter is virtually an 1-to-1 mapping.}
\end{figure}

\clearpage
\paragraph*{S1 Table.}
\label{S1_Table}
Benchmarking different methods using $\mathtt{\addnew{lrscaf}}$, \npscarf{}, \npgraph{}, $\mathtt{hybridSPAdes}$ and \unicycler{} hybrid assembler with the synthetic data set.

@@

& \addnew{lrscaf} & \addnew{-} & \addnew{-} & \addnew{-} & \addnew{-} & \addnew{-} & \addnew{-} \\*

@@

& \addnew{lrscaf} & \addnew{4113104} & \addnew{6} & \addnew{4002298} & \addnew{0} & \addnew{0.00} & \addnew{0.00}\\*

@@

& \addnew{lrscaf} & \addnew{4119020} & \addnew{12} & \addnew{4002117} & \addnew{0} & \addnew{0.03} & \addnew{0.79}\\* 

@@

& \addnew{lrscaf} & \addnew{3896799} & \addnew{4} & \addnew{3886381} & \addnew{0} & \addnew{0.47} & \addnew{
0.36}\\*

@@

& \addnew{lrscaf} & \addnew{4367657} & \addnew{27} & \addnew{1035967} & \addnew{13} & \addnew{30.11} & \addnew{1.42}\\* 

@@

& \addnew{lrscaf} & \addnew{4732001} & \addnew{16} & \addnew{692794} & \addnew{7} & \addnew{3.42} & \addnew{0.58}\\*

@@

& \addnew{lrscaf} & \addnew{5410854} & \addnew{21} & \addnew{910949} & \addnew{9} & \addnew{13.70} & \addnew{0.87}\\*

@@

& \addnew{lrscaf} & \addnew{5512518} & \addnew{20} & \addnew{777919} & \addnew{26} & \addnew{3.52} & \addnew{0.38}\\*

@@

& \addnew{lrscaf} & \addnew{5823340} & \addnew{38} & \addnew{559414} & \addnew{7} & \addnew{5.92} & \addnew{1.27}\\*

@@

& \addnew{lrscaf} & \addnew{5470282} & \addnew{8} & \addnew{2951488} & \addnew{9} & \addnew{3.04} & \addnew{2.62}\\*

@@

& \addnew{lrscaf} & \addnew{4260869} & \addnew{12} & \addnew{3215170} & \addnew{8} & \addnew{6.69} & \addnew{2.27}\\*

@@

& \addnew{lrscaf} & \addnew{12486884} & \addnew{214} & \addnew{254587} & \addnew{27} & \addnew{16.45} & \addnew{1.97}\\*

@@

& \addnew{lrscaf} & \addnew{4678609} & \addnew{175} & \addnew{376897} & \addnew{26} & \addnew{3.91} & \addnew{2.35}\\*

@@

& \addnew{lrscaf} & \addnew{5266645} & \addnew{128} & \addnew{429353} & \addnew{27} & \addnew{9.72} & \addnew{1.19}\\*

@@

& \addnew{lrscaf} & \addnew{2231488} & \addnew{10} & \addnew{683404} & \addnew{6} & \addnew{11.96} & \addnew{1.37}\\*

>>>>

%% END MACROS SECTION
\usepackage[normalem]{ulem}
%% npGraph
\newcommand{\npscarf}{$\mathtt{npScarf}$}
\newcommand{\npscarfg}{$\mathtt{npScarf\_wag}$}
\newcommand{\npreader}{$\mathtt{npReader}$}
\newcommand{\npanalysis}{$\mathtt{npAnalysis}$}
\newcommand{\npbarcode}{$\mathtt{npBarcode}$}
\newcommand{\npgraph}{$\mathtt{npGraph}$}
\newcommand{\lrscaf}{$\mathtt{npGraph}$}
\newcommand{\canu}{$\mathtt{Canu}$}
\newcommand{\unicycler}{$\mathtt{Unicycler}$}
\newcommand{\spades}{$\mathtt{SPAdes}$}
\newcommand{\hspades}{$\mathtt{hybridSPAdes}$}
\newcommand{\albacore}{$\mathtt{Albacore}$}
\newcommand{\racon}{$\mathtt{Racon}$}
\newcommand{\metrichor}{$\mathtt{Metrichor}$}
\newcommand{\minimap}{$\mathtt{minimap2}$}
\newcommand{\miniasm}{$\mathtt{miniasm}$}
\newcommand{\bwa}{$\mathtt{BWA\text{-}MEM}$}
\newcommand{\lrscarf}{$\mathtt{LRScaf}$}

@@

One particular strength of ONT technology is the production of ultra long reads. 
This is complementary to the dominant short read sequencing technology Illumina which is cheaper and has higher per base read quality but is unable to resolve the complex regions of the genome due to its read length limitation. 
Hybrid assembly with ONT reads provides the opportunity to use long reads to fully resolve existing microbial genome assemblies. However, in order achieve this in an effective manner it is necessary to avoid both under-sequencing (and thus not completely resolving the assembly) or over-sequencing (and thus incurring unnecessary costs and time-to -results).  In addition, the release of Read Until API provides the opportunity to selectively enrich parts of the genome, as has been implemented in customised targeted sequencing applications ~\cite{Payne2020}.   The combination of ReadUntil with  streaming hybrid assembly opens the possibility for further efficiency gains by targeting sequencing to unresolved regions of the genome.

@@

The first step is to cluster the big anchors (longer than $10$Kbp and in/out degree less than 2) based on their \emph{kmer} coverage.
To achieve this, we applied DBSCAN clustering algorithm \cite{Ester96adensity-based}, which utilises a customised metric function to map contigs into a one-dimensional space.  In order to 
 define a customised metric which is simple and fast to calculate, we assumed that a single long contig itself consists of a Poisson distribution of \emph{k-mers} count with the mean approximated by the contig's coverage. 
%The rationale is to approximate a coverage value of an extremely long contig (which can be split into more than 10,000 \emph{k-mers}) with a sampled mean of a Poisson distribution (of \emph{k-mers} count). 
The metric is then determined by a distance function of two Poisson distributions based on Kullback-Leibler divergence (or relative entropy) between the Poisson distribution representing each contig\cite{Kullback1951information}.

@@

Due to false positive alignments from shorter contigs to the long reads, not all of the reported step nodes would appear in the ultimate path resolved by the bridge. 
In most cases, the accumulated score of each step indicates its likelihood to be the true component of the final solution.
For that reason, a strategy similar to binary search is employed to find a path across 2 anchors of a bridge as shown in Algorithm~\ref{algo:binsearch}. Briefly, it recursively looks for sub-paths connecting pairs of step nodes in-between 2 anchors and return a final continuous path with the highest score.
We define Algorithm~\ref{algo:findpath} as the \emph{base} sub-rountine to find paths between a \emph{connectable} pair of two step nodes given their estimated distance. In which, function 
\[
\mathtt{shortestTree}(\overrightarrow{vertex},distance) : (V,Z) \rightarrow V^n
\]
from line 3 builds a shortest tree rooted from $\overrightarrow{v}$, following its direction until a distance of approximately $d$ (with a tolerance for the nanopore error rate) is reached. This task is implemented based on Dijkstra algorithm.
This tree is used on line 4 and in function $includedIn()$ on line 19 to filter out any node or edge with ending nodes that do not belong to the tree.

@@

% Results and Discussion can be combined.
\section*{Results}
\subsection*{Evaluation using synthetic data}
To evaluate the performance of the method, \npgraph{} 1.1 was tested along with \spades{}, \spades{} hybrid from version 3.13.1, ~\cite{AntipovKM2015}, \npscarf{} (japsa 1.7-02), \lrscaf{} version 1.1.9 and \unicycler{} version 0.4.6 using \unicycler{}`s synthetic data set~ \cite{Wick2017unicycler} . 
The data set is a simulation of Illumina and MinION raw data, generated \emph{in silico} based on available microbial references.
We ran hybrid assembly methods using the entire nanopore data and the reciprocal results were evaluated by QUAST 5.0.2~\cite{Mikheenko2018quast5}.

@@

\lrscaf{} has been designed as a computationally efficient hybrid assembly tool which is scalable to large genomes.  We ran \lrscaf{} using default parameters on the extended benchmarking set presented in Supplementary table 1.  We observed that  \lrscaf{} assemblies had fewer misassemblies than \npscarf{}, and also exhibited a low mismatch and indel rate, but produced more fragmented assemblies than either \npgraph{} or \unicycler{}.

@@

\begin{figure}[!ht]
 \includegraphics[width=\textwidth]{images/FigS2.eps}
 \caption{Alignments of an \emph{Enterobacter~cloacae} reference genome to assembly sequences generated by  (a)~\npgraph{} and (b)~\unicycler{}. The former suggests a structural variant, the latter is virtually an 1-to-1 mapping.}
\end{figure}

\clearpage
\paragraph*{S1 Table.}
\label{S1_Table}
Benchmarking different methods using \lrscaf{}, \npscarf{}, \npgraph{}, $\mathtt{hybridSPAdes}$ and \unicycler{} hybrid assembler with the synthetic data set.

@@

& LRScaf & - & - & - & - & - & - \\*

@@

& LRScaf & 4113104 & 6 & 4002298 & 0 & 0.00 & 0.00\\*

@@

& LRScaf & 4119020 & 12 & 4002117 & 0 & 0.03 & 0.79\\* 

@@

& LRScaf & 3896799 & 4 & 3886381 & 0 & 0.47 & 0.36\\*

@@

& LRScaf & 4367657 & 27 & 1035967 & 13 & 30.11 & 1.42\\* 

@@

& LRScaf & 4732001 & 16 & 692794 & 7 & 3.42 & 0.58\\*

@@

& LRScaf & 5410854 & 21 & 910949 & 9 & 13.70 & 0.87\\*

@@

& LRScaf & 5512518 & 20 & 777919 & 26 & 3.52 & 0.38\\*

@@

& LRScaf & 5823340 & 38 & 559414 & 7 & 5.92 & 1.27\\*

@@

& LRScaf & 5470282 & 8 & 2951488 & 9 & 3.04 & 2.62\\*

@@

& LRScaf & 4260869 & 12 & 3215170 & 8 & 6.69 & 2.27\\*

@@

& LRScaf & 12486884 & 214 & 254587 & 27 & 16.45 & 1.97\\*

@@

& LRScaf & 4678609 & 175 & 376897 & 26 & 3.91 & 2.35\\*

@@

& LRScaf & 5266645 & 128 & 429353 & 27 & 9.72 & 1.19\\*

@@

& LRScaf & 2231488 & 10 & 683404 & 6 & 11.96 & 1.37\\*

### 2020-09-08 plos_latex_edit.tex
Lachlan revision 1

%% END MACROS SECTION
\usepackage[normalem]{ulem}
\newcommand\cancel[1]{\sout{#1}}
\newcommand\addnew[1]{\textcolor{red}{#1}}
%% npGraph
\newcommand{\npscarf}{$\mathtt{npScarf}$}
\newcommand{\npscarfg}{$\mathtt{npScarf\_wag}$}
\newcommand{\npreader}{$\mathtt{npReader}$}
\newcommand{\npanalysis}{$\mathtt{npAnalysis}$}
\newcommand{\npbarcode}{$\mathtt{npBarcode}$}
\newcommand{\npgraph}{$\mathtt{npGraph}$}
\newcommand{\canu}{$\mathtt{Canu}$}
\newcommand{\unicycler}{$\mathtt{Unicycler}$}
\newcommand{\spades}{$\mathtt{SPAdes}$}
\newcommand{\hspades}{$\mathtt{hybridSPAdes}$}
\newcommand{\albacore}{$\mathtt{Albacore}$}
\newcommand{\racon}{$\mathtt{Racon}$}
\newcommand{\metrichor}{$\mathtt{Metrichor}$}
\newcommand{\minimap}{$\mathtt{minimap2}$}
\newcommand{\miniasm}{$\mathtt{miniasm}$}
\newcommand{\bwa}{$\mathtt{BWA\text{-}MEM}$}

@@

% Results and Discussion can be combined.
\section*{Results}
\subsection*{Evaluation using synthetic data}
To evaluate the performance of the method, \npgraph{} 1.1 was tested along with \spades{}, \spades{} hybrid from version 3.13.1, ~\cite{AntipovKM2015}, \npscarf{} (japsa 1.7-02), and \unicycler{} version 0.4.6 using \unicycler{}`s synthetic data set~ \cite{Wick2017unicycler} . 
\addnew{}
The data set is a simulation of Illumina and MinION raw data, generated \emph{in silico} based on available microbial references.
We ran hybrid assembly methods using the entire nanopore data and the reciprocal results were evaluated by QUAST 5.0.2~\cite{Mikheenko2018quast5}.

@@

>>>>

%% END MACROS SECTION
\usepackage[normalem]{ulem}
\newcommand\cancel[1]{\sout{#1}}
\newcommand\addnew[1]{\textcolor{red}{#1}}
%% npGraph
\newcommand{\npscarf}{$\mathtt{npScarf}$}
\newcommand{\npscarfg}{$\mathtt{npScarf\_wag}$}
\newcommand{\npreader}{$\mathtt{npReader}$}
\newcommand{\npanalysis}{$\mathtt{npAnalysis}$}
\newcommand{\npbarcode}{$\mathtt{npBarcode}$}
\newcommand{\npgraph}{$\mathtt{npGraph}$}
\newcommand{\lrscaf}{$\mathtt{npGraph}$}
\newcommand{\canu}{$\mathtt{Canu}$}
\newcommand{\unicycler}{$\mathtt{Unicycler}$}
\newcommand{\spades}{$\mathtt{SPAdes}$}
\newcommand{\hspades}{$\mathtt{hybridSPAdes}$}
\newcommand{\albacore}{$\mathtt{Albacore}$}
\newcommand{\racon}{$\mathtt{Racon}$}
\newcommand{\metrichor}{$\mathtt{Metrichor}$}
\newcommand{\minimap}{$\mathtt{minimap2}$}
\newcommand{\miniasm}{$\mathtt{miniasm}$}
\newcommand{\bwa}{$\mathtt{BWA\text{-}MEM}$}
\newcommand{\lrscarf}{$\mathtt{LRScaf}$}

@@

% Results and Discussion can be combined.
\section*{Results}
\subsection*{Evaluation using synthetic data}
To evaluate the performance of the method, \npgraph{} 1.1 was tested along with \spades{}, \spades{} hybrid from version 3.13.1, ~\cite{AntipovKM2015}, \npscarf{} (japsa 1.7-02), \addnew{\lrscaf{} version 1.1.9} and \unicycler{} version 0.4.6 using \unicycler{}`s synthetic data set~ \cite{Wick2017unicycler} . 
\addnew{}
The data set is a simulation of Illumina and MinION raw data, generated \emph{in silico} based on available microbial references.
We ran hybrid assembly methods using the entire nanopore data and the reciprocal results were evaluated by QUAST 5.0.2~\cite{Mikheenko2018quast5}.

@@

\addnew{\lrscaf{} has been designed as a computationally efficient hybrid assembly tool which is scalable to large genomes.  We ran \lrscaf{} using default parameters on the extended benchmarking set presented in Supplementary table 1.  We observed that  \lrscaf{} assemblies had fewer misassemblies than \npscarf{}, and also exhibited a low mismatch and indel rate, but produced more fragmented assemblies than either \npgraph{} or \unicycler{}. }
